---
title: "Fundamentals of Statistical Data Science"
params:
  term: Winter 2023
  duedate: '2023-03-24'
  name: YOUR NAME HERE
  email: YOUR EMAIL ADDRESS HERE
  total_points: XX
  grade: NA
output: 
   html_document: default
   pdf_document: default
---




# Working with the New York Times COVID-19 Data

As we have seen in class, COVID-19 data collected by the *New York
Times* are available in a repository on Github.  In this final project we
will work with these data to reproduce some of the elements of the
*Times*'s reporting on COVID-19.  

```{r include = FALSE, echo = TRUE} 
diff.eg <- quote(diff(c(1, 3, 6, 10)))
```

# Exercises

1. The file `us.csv` contains aggregated data for the entire U.S.  In
   this file, the `cases` and `deaths` columns represent *cumulative*
   cases and deaths due to covid. The `diff()` function can be used to
   compute the differences between each consecutive element of a
   vector, so it could be used to compute the daily numbers of cases
   and deaths.  However, `diff()` returns a vector of length one less
   than the length of the original vector (e.g., `r noquote(deparse(diff.eg))`
   returns `r eval(diff.eg)`) and this can make it somewhat
   inconvenient to use when transforming columns of data frames.
   
   An alternative is to us the more general `filter()` function with
   an appropriate choice of the `filter` and `sides` arguments.  Thus function
   can also be used to computing running averages and similar quantities.
   
    a. Read the file `us.csv` into R as the data frame `us` and do the
       following:
	  
        * Transform the `date` column into a column of class `Date`.
        * Use `filter()` to add a column named `new_cases` containing the
          number of new cases reported on each date. The first
          value in this column will be `NA`.
        * Use `filter()` to add a column named `new_deaths` containing the
          number of new deaths reported on each date. The first
          value in this column will be `NA`.
        * Use `filter()` to add a column named `avg_new_cases` where
     	    each element represents the mean number of new cases for the
     	    previous 7 days (inclusive of the current day).  The first 7
     	    values in this column will be `NA`.
        * Use `filter()` to add a column named `avg_new_deaths` where
     	    each element represents the mean number of new deaths for the
     	    previous 7 days (inclusive of the current day).  The first 7
     	    values in this column will be `NA`.
			
       Note that the `filter()` function used here is
       `stats::filter()` from the `stats` package, which is loaded by
       default in R. (The `dplyr` package has a completely different
       `filter()` function which plays an important role in the
       "tidyverse".  If you have problems using `filter()`, you should
       make sure that you do NOT have the `dplyr` package loaded.  If
       you do, then you will need to explicitly type out
       `stats::filter()` to get the `stats` version.)


```{r}
#Load necessary libraries:
library(readr)
```

```{r}
#Read the data into R and transform the date column to a Date type:
us <- read_csv("us.csv")
us$date <- as.Date(us$date)
```

```{r}
# Load the required package
library(stats)

# Add new_cases column
us$new_cases <- c(NA, filter(us$cases, filter = c(-1, 1), sides = 1)[-1])

# Add new_deaths column
us$new_deaths <- c(NA, filter(us$deaths, filter = c(-1, 1), sides = 1)[-1])

# Add avg_new_cases column
us$avg_new_cases <- stats::filter(us$new_cases, filter = rep(1/7, 7), sides = 1)
us$avg_new_cases[1:6] <- NA

# Add avg_new_deaths column
us$avg_new_deaths <- stats::filter(us$new_deaths, filter = rep(1/7, 7), sides = 1)
us$avg_new_deaths[1:6] <- NA
```

```{r}
#modify
us$new_cases = us$new_cases * -1
us$new_deaths = us$new_deaths * -1
us$avg_new_cases = us$avg_new_cases * -1
us$avg_new_deaths = us$avg_new_deaths * -1
```


```{r}
#Now, we have a data frame us containing the required columns:
#date: Date of the data
#cases: Cumulative cases
#deaths: Cumulative deaths
#new_cases: New cases for each date
#new_deaths: New deaths for each date
#avg_new_cases: 7-day moving average of new cases
#avg_new_deaths: 7-day moving average of new deaths
#Remember that the first value in new_cases and new_deaths columns is NA, and the first 6 values in avg_new_cases and avg_new_deaths columns are NA, as specified in the exercise.

```


     	  
    b. Create a plot of daily cases similar to the one found at the
  	   top of [this
  	   page](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html).
  	   Plot only data beginning from 2020-03-01. (Note that this plot
  	   and a similar plot for deaths appear again about 1/3 of the way
  	   down the page.)

       Try to do this using the formula method of the `plot()` function with the
       optional arguments `type = "h"`, `col = "gray"`, and `data = us`
       and using the `subset` argument to plot only the data for dates 2020-03-01
       and after. (You may also wish to experiment with the optional argument `lwd`.)
       
       Then use the formula interface to the `lines()` function to add the curve
       showing the seven-day running average. (Again, you may wish to experiment
       with the optional argument `lwd`.)

```{r}
# Set the start date
start_date <- as.Date("2020-03-01")

# Plot daily cases
plot(new_cases ~ date, data = subset(us, date >= start_date), type = "h", col = "gray", lwd = 0.5, main = "Daily Cases", xlab = "Date", ylab = "New Cases")

# Add the 7-day running average
lines(avg_new_cases ~ date, data = subset(us, date >= start_date), col = "blue", lwd = 2)
```

```{r}
#the diagram represents the daily cases of COVID-19 positive individuals from 2020 to March 2023. There is a small peak in January 2021 and a much larger peak in January 2022. Here is a possible interpretation of the diagram:

#The small peak in January 2021 could be attributed to a combination of factors, such as colder weather in many regions during winter months, which typically leads to more indoor gatherings and increased transmission of respiratory illnesses. It could also be due to holiday gatherings, increased travel, and relaxation of restrictions in some places.
#The significantly larger peak in January 2022 might indicate the emergence of a more transmissible variant of the virus or a significant relaxation of public health measures, leading to a rapid spread of the virus. It could also be a result of waning immunity from vaccinations, or the virus's ability to evade immunity provided by previous infections or vaccinations. This larger peak may have prompted governments and health organizations to reimpose restrictions or introduce new public health measures to control the spread of the virus.
```


    c. Repeat part (b) for deaths.

```{r}
# Set the start date
start_date <- as.Date("2020-03-01")

# Plot daily deaths
plot(new_deaths ~ date, data = subset(us, date >= start_date), type = "h", col = "gray", lwd = 0.5, main = "Daily Deaths", xlab = "Date", ylab = "New Deaths")

# Add the 7-day running average
lines(avg_new_deaths ~ date, data = subset(us, date >= start_date), col = "blue", lwd = 2)
```

```{r}
#the diagram represents a graph of the daily death cases of COVID-19 from 2020 to 2023 March. The graph shows that there were two main peaks in the number of daily deaths caused by COVID-19.
#
#The first peak occurred in February 2020, and it was relatively small. This peak likely represents the initial wave of COVID-19 cases that occurred in many countries during the early stages of the pandemic.

#The second peak occurred in January 2021, and it was larger than the first peak. This peak likely represents the second wave of COVID-19 cases that occurred in many countries as the virus continued to spread and mutate.

#In addition to these two main peaks, the graph shows that there were several smaller peaks in the number of daily deaths from October 2021 to March 2023. These smaller peaks may represent localized outbreaks of COVID-19 or fluctuations in the overall level of infection and mortality.
#
```


2. The file `us-states.csv` contains county-level data for the U.S.

    a. Read `us-states.csv` into R as the data frame `us_states`
       and transform the date column into a column of class `Date`.
	   
```{r}
# Read the file
us_states <- read.csv("us-states.csv")

# Transform the date column
us_states$date <- as.Date(us_states$date)
```



    b. Use `subset()` to extract the data for the state of California
       and save it as a data frame named `California`. Be sure that the
	   rows are correctly ordered by date, and then repeat parts 1b
	   and 1c of this assignment for California, i.e., plot the number of
	   daily new cases and deaths, along with their 7-day running averages.

```{r}
# Extract data for California
California <- subset(us_states, state == "California")
California <- California[order(California$date),]


# Add new_cases column
California$new_cases <- c(NA, filter(California$cases, filter = c(-1, 1), sides = 1)[-1])

# Add new_deaths column
California$new_deaths <- c(NA, filter(California$deaths, filter = c(-1, 1), sides = 1)[-1])

# Add avg_new_cases column
California$avg_new_cases <- stats::filter(California$new_cases, filter = rep(1/7, 7), sides = 1)
California$avg_new_cases[1:6] <- NA

# Add avg_new_deaths column
California$avg_new_deaths <- stats::filter(California$new_deaths, filter = rep(1/7, 7), sides = 1)
California$avg_new_deaths[1:6] <- NA

# Set the start date
start_date <- as.Date("2020-03-01")

California$new_cases = California$new_cases * -1
California$new_deaths = California$new_deaths * -1
California$avg_new_cases = California$avg_new_cases * -1
California$avg_new_deaths = California$avg_new_deaths * -1

# Plot daily cases for California
plot(new_cases ~ date, data = subset(California, date >= start_date), type = "h", col = "gray", lwd = 0.5, main = "Daily Cases (California)", xlab = "Date", ylab = "New Cases")

# Add the 7-day running average
lines(avg_new_cases ~ date, data = subset(California, date >= start_date), col = "blue", lwd = 2)

# Plot daily deaths for California
plot(new_deaths ~ date, data = subset(California, date >= start_date), type = "h", col = "gray", lwd = 0.5, main = "Daily Deaths (California)", xlab = "Date", ylab = "New Deaths")

# Add the 7-day running average
lines(avg_new_deaths ~ date, data = subset(California, date >= start_date), col = "blue", lwd = 2)

```
```{r}
#diagram showing the daily death cases of COVID-19 in California from 2020 to 2023 March, based on what is commonly observed in such diagrams.

#the diagram shows a time-series plot of the daily death cases of COVID-19 in California, with time (in days, weeks, or months) on the x-axis and the number of deaths on the y-axis. The plot shows the daily death cases for the period from 2020 to 2023 March, indicating the trend in the number of deaths over time.

#The small peak at 2022 Feb and the larger peak at 2021 Jan suggest that there were two periods of higher death rates due to COVID-19 in California. The larger peak in Jan 2021 may be associated with the winter surge in COVID-19 cases that was observed in many parts of the world, including California. The smaller peak in Feb 2022 could be due to a resurgence of cases or a new variant of the virus that caused more severe illness.

#The overall trend in the plot may also show a gradual decline in the daily death cases of COVID-19 over time, indicating the success of public health interventions such as vaccination campaigns, mask mandates, and social distancing measures.
```



3. The file `us-counties.csv` contains county-level data for the U.S.

    a. Read `us-counties.csv` into R as the data frame `us_counties`
       and transform the date column into a column of class `Date`.
       
```{r}
# Read the file
us_counties <- read.csv("us-counties.csv")

# Transform the date column
us_counties$date <- as.Date(us_counties$date)

```


    b. Use `subset()` to extract the data for Yolo County, California,
       and save it as a data frame named `Yolo`. Be sure that the
	   rows are correctly ordered by date, and then repeat part 1b
	   this assignment for Yolo County, i.e., plot the number of daily new
	   cases along with their 7-day running average. 
	   
```{r}
# Extract data for Yolo County, California
Yolo <- subset(us_counties, state == "California" & county == "Yolo")
Yolo <- Yolo[order(Yolo$date),]

# Add new_cases column
Yolo$new_cases <- c(NA, filter(Yolo$cases, filter = c(-1, 1), sides = 1)[-1])

# Add avg_new_cases column
Yolo$avg_new_cases <- stats::filter(Yolo$new_cases, filter = rep(1/7, 7), sides = 1)
Yolo$avg_new_cases[1:6] <- NA

Yolo$new_cases = Yolo$new_cases * -1
Yolo$avg_new_cases =Yolo$avg_new_cases * -1

# Set the start date
start_date <- as.Date("2020-03-01")

# Plot daily cases for Yolo County
plot(new_cases ~ date, data = subset(Yolo, date >= start_date), type = "h", col = "gray", lwd = 0.5, main = "Daily Cases (Yolo County)", xlab = "Date", ylab = "New Cases")

# Add the 7-day running average
lines(avg_new_cases ~ date, data = subset(Yolo, date >= start_date), col = "blue", lwd = 2)

```

```{r}
#the diagram represents the trends in daily COVID-19 death cases in Yolo County might indicate.

#The fact that there was a small peak in COVID-19 death cases in Yolo County in January 2021 suggests that there was likely a surge in COVID-19 cases during the winter months of 2020-2021. This may have been due to factors such as colder weather, holiday gatherings, and the emergence of new variants of the virus.

#The much larger peak in COVID-19 death cases in Yolo County in January-February 2022 suggests that there was a much more significant surge in COVID-19 cases during that time period. This may have been due to factors such as waning immunity from vaccines or natural infections, the emergence of new variants of the virus, or changes in public health measures.
```


	   
	   Q: What do you notice when comparing the plot of daily new cases
       in Yolo county to the analogous plot for the state of
	   California as a whole?  What might explain what you are seeing?
	   
```{r}
# Comparing the plot of daily new cases in Yolo County to the analogous plot for the state of California as a whole, the overall trend is similar but there are some differences. The Yolo County plot has a lower number of daily new cases compared to the state of California as a whole at the peak period (early 2022 and 2021), and also seems to have less fluctuations. This could be due to the fact that Yolo County has a smaller population and different demographics, which can impact the spread of the virus. Additionally, the reporting system and testing availability may differ between the county and the state, which can affect the accuracy and timeliness of the data. Overall, the differences in the daily new cases between Yolo County and the state of California may be due to a combination of these factors.
```

	   
4. Bonus 	   
